Using occupancy theory and Bayesian models to estimate otter populations in southeastern Minnesota

### 2010-03-04 Presentation/Presentation.tex
Hopefully fixed the crash-happy convergence graphic.
Edited baysian slide
Changed aggregation slide
A few more changes
poop

\begin{frame}{Estimating occupancy}
	\begin{itemize}
		\item Ecologists often conduct surveys to find the occupancy rate
		\item Not observing the species can mean two things:
		\begin{itemize}
			\item The species does not occupy the area
			\item The species went undetected during surveying
		\end{itemize}
		\item Need to be able to estimate probability of observing the species
		given its presence
	\end{itemize}
\end{frame}

@@

\begin{frame}{Types of occupancy models}
	\begin{itemize}
		\item Single species, single season
		\item Na\"ive occupancy: proportion of sites found to be occupied
		\item Hierarchical models
	\end{itemize}
	\begin{center}
		\resizebox{7cm}{!}{\includegraphics{../SimpleHierarchicalModel.pdf}}
	\end{center}
\end{frame}

\begin{frame}{Parameters}
	\begin{itemize}
		\item $\psi=$ The probability a track is occupied
		\item $\theta=$ The probability a track is laid down in a site on a
		given day, given that the site is occupied
		\item $p= $ The probability an observer sees a track given that a track
		has been made
	\end{itemize}
	\begin{center}
		\resizebox{7cm}{!}{\includegraphics{../SimpleModelWithParams.pdf}}
	\end{center}
\end{frame}

@@

\begin{frame}{Bayesian vs. likelihood occupancy models}
	\begin{itemize}
		\item Maximum likelihood assumes random sample from large population
		\begin{itemize}
			\item Entire section of river was sampled
		\end{itemize}
		\item Most occupancy models have two random processes -- occupancy and
		detection
		\begin{itemize}
			\item We added false detection and track-laying
			\item Extra parameters made likelihood model complicated
		\end{itemize}
	\end{itemize}
\end{frame}

@@

\begin{frame}{Convergence}
	\begin{center}
		\resizebox{9cm}{!}{\includegraphics{historyConvergence.pdf}} \\
		\includegraphics[scale=0.8]{distnConvergence.pdf} \\
	\end{center}
\end{frame}

\begin{frame}{Convergence}
	\begin{center}
		\resizebox{9cm}{!}{\includegraphics{historyNoConvergence.pdf}} \\
		\includegraphics[scale=0.8]{distnNoConvergence.pdf} \\
	\end{center}
\end{frame}

\begin{frame}{Issue with false detection}
	\begin{itemize}
		\item $(p=x, E=y, \psi=z)$ equivalent in terms of likelihood to
		$(p=y,E=x,\psi=1-z)$.
		\item Leads to bimodal posterior distributions for $p$ and $E$.
		\item Solution: restrict $p$ to interval $(0.6,1)$, $E$ to interval
		$(0,0.4)$.
	\end{itemize}
\end{frame}

@@

\begin{frame}{Testing the CAR model}
	\begin{itemize}
		\item Simulated data with correlation in $\theta$
		\begin{itemize}
			\item $P(\text{track in site }i|\text{no track in site }i-1)=\theta$
			\item $P(\text{track in site }i|\text{track in site }i-1)=\theta+
			\alpha$
		\end{itemize}
		\item CAR model did not converge
		\item Simulated data more independent than actual data (from $S$)
		\item Added similar correlation in $\psi$
		\begin{itemize}
			\item CAR model converged more often
			\item Independence more similar to that of actual data
		\end{itemize}
		\item When CAR model converged, estimates from CAR and standard models
		similar for the same data
	\end{itemize}
\end{frame}

\section{Discussion}
\begin{frame}{Obtaining convergence on the data}
	\begin{itemize}
		\item All of 2003 data converged with standard model
		\item Nothing in 2004 converged
		\begin{itemize}
			\item Removed false detection, 402m in 2004 converged
		\end{itemize}
		\item CAR model never converged with $E$. Without $E$, poor
		estimates for $\psi$
	\end{itemize}
\end{frame}

@@

\begin{frame}{Summary and interpretation of estimates}
	\begin{center}
		\includegraphics[scale=0.5]{Aggregation.pdf}
	\end{center}
	\begin{itemize}
		\item Estimates for all parameters increase with plot size
		\item Possible hypotheses:
		\begin{itemize}
			\item \(\psi\) and \(\theta\)
			\item \(p\) and \(E\)
		\end{itemize}
		\item Used data from all plots, not alternating plots
		\begin{itemize}
			\item More data more useful than more independence
		\end{itemize}
	\end{itemize}
\end{frame}

>>>>

\begin{frame}{Estimating occupancy}
	\begin{itemize}
		\item Ecologists often conduct surveys to find the occupancy rate
		\item Na\"ive occupancy: proportion of sites found to be occupied
		\item Not observing the species can mean two things:
		\begin{itemize}
			\item The species does not occupy the area
			\item The species went undetected during surveying
		\end{itemize}
		\item Need to be able to estimate probability of observing the species
		given its presence
	\end{itemize}
\end{frame}

@@

\begin{frame}{Types of occupancy models}
	\begin{itemize}
		\item Single species, single season
		\item Hierarchical models
	\end{itemize}
	\begin{center}
		\resizebox{7cm}{!}{\includegraphics{../SimpleHierarchicalModel.pdf}}
	\end{center}
\end{frame}

\begin{frame}{Parameters}
	\begin{itemize}
		\item $\psi=$ The probability a site is occupied
		\item $\theta=$ The probability a track is laid down in a site on a
		given day, given that the site is occupied
		\item $p= $ The probability an observer sees a track given that a track
		has been made
	\end{itemize}
	\begin{center}
		\resizebox{7cm}{!}{\includegraphics{../SimpleModelWithParams.pdf}}
	\end{center}
\end{frame}

@@

\begin{frame}{Bayesian vs. likelihood occupancy models}
	\begin{itemize}
		\item Maximum likelihood assumes random sample from large population
		\begin{itemize}
			\item Entire section of river was sampled
		\end{itemize}
		\item Most occupancy models have two random processes -- occupancy and
		detection
		\begin{itemize}
			\item We added false detection and track-laying
			\item Extra parameters made likelihood model complicated
		\end{itemize}
		\item Spatial correlation is easier with a Bayesian model
	\end{itemize}
\end{frame}

@@

\begin{frame}{Convergence}
	\begin{center}
		\resizebox{8.5cm}{!}{\includegraphics{historyConvergence.pdf}} \\
		\includegraphics[scale=0.8]{distnConvergence.pdf} \\
	\end{center}
\end{frame}

\begin{frame}{Convergence}
	\begin{center}
		\resizebox{8.5cm}{!}{\includegraphics{historyNoConvergence.pdf}} \\
		\includegraphics[scale=0.8]{distnNoConvergence.pdf} \\
	\end{center}
\end{frame}

\begin{frame}{Issue with false detection}
	\begin{itemize}
		\item $(p=x, E=y, \psi=z)$ equivalent in terms of likelihood to
		$(p=y,E=x,\psi=1-z)$.
		\item Leads to bimodal posterior distributions.
		\item Solution: restrict $p$ to interval $(0.6,1)$, $E$ to interval
		$(0,0.4)$.
	\end{itemize}
\end{frame}

@@

\begin{frame}{Testing the CAR model}
	\begin{itemize}
		\item Simulated data with correlation in $\theta$
		\begin{itemize}
			\item $P(\text{track in site }i|\text{no track in site }i-1)=\theta$
			\item $P(\text{track in site }i|\text{track in site }i-1)=\theta+
			\alpha$
		\end{itemize}
		\item CAR model did not converge
		\item Simulated data less dependent than actual data (from $S$)
		\item Added similar correlation in $\psi$
		\begin{itemize}
			\item CAR model converged more often
			\item Dependence more similar to that of actual data
		\end{itemize}
		\item When CAR model converged, estimates from CAR and standard models
		similar for the same data
	\end{itemize}
\end{frame}

\section{Discussion}
\begin{frame}{Obtaining convergence on the data}
	\begin{itemize}
		\item All of 2003 data converged with standard model
		\item Nothing in 2004 converged
		\begin{itemize}
			\item Removed false detection
			\item 402m in 2004 converged, but estimates not dependable
		\end{itemize}
		\item CAR model never converged with $E$. Without $E$, poor
		estimates for $\psi$
	\end{itemize}
\end{frame}

@@

\begin{frame}{Summary and interpretation of estimates}
	\begin{itemize}
		\item Estimates for all parameters increase with plot size
		\item Possible hypotheses:
		\begin{itemize}
			\item \(\psi\) and \(\theta\)
			\item \(p\) and \(E\)
		\end{itemize}
		\item Used data from all plots, not alternating plots
		\begin{itemize}
			\item More data more useful than more independence
		\end{itemize}
	\end{itemize}
	\begin{center}
		\includegraphics[scale=0.5]{Aggregation.pdf}
	\end{center}
\end{frame}
